---
title: "Daily New Cases of COVID-19 in Danish Municipalities"
author: Tommi Suvitaival, tommi.raimo.leo.suvitaival@regionh.dk, Steno Diabetes Center
  Copenhagen
date: "`r Sys.Date()`"
output: github_document
---

# Introduction

* Municipality-level data of daily new cases of COVID-19 in Denmark
* Data from Statens Serum Institut
  + at https://www.ssi.dk/sygdomme-beredskab-og-forskning/sygdomsovervaagning/c/covid19-overvaagning
  + behind  "fil med overvågningsdata"

# Load Data

```{r}

# Data from Statens Serum Institut
# at https://www.ssi.dk/sygdomme-beredskab-og-forskning/sygdomsovervaagning/c/covid19-overvaagning
# behind "fil med overvågningsdata"
# (Direct link:
#  https://files.ssi.dk/Data-Epidemiologiske-Rapport-10092020-77gf ;
#  accessed 11.9.2020.
# )

data.loaded <-
  readr::read_delim(
    file = "H:/projects/covid19/data/SSI/Municipality_cases_time_series.csv", 
    delim = ";",
    escape_double = FALSE,
    trim_ws = TRUE,
    col_names = TRUE
  )

```

# Prepare Data

```{r}

data <- data.loaded

data$"date_sample" <- lubridate::as_date( x = data$"date_sample" )

```

# Proportion of Municipalities with Daily New COVID-19 Cases

## Compute Values

```{r}

data.plot <- data

data.plot$"Pcnt_municipalities_with_new_cases" <- 
  rowMeans( x = data[ , -1 ] > 0 ) * 100

data.plot$"Total_new_cases" <- rowSums( x = data[ , -1 ] )

```

```{r}

breaks.date <-
  ggplot2::scale_x_date(
    breaks = as.Date( paste0( "2020-0", 3:9, "-01" ) ),
    date_labels = "%b"
  )

```

## Plot

```{r Cases-v-Date, dpi = 600, dev = c( "png" )}

ggplot2::ggplot(
  data = data.plot,
  mapping =
    ggplot2::aes(
      x = date_sample,
      y = Pcnt_municipalities_with_new_cases
    )
) +
  ggplot2::geom_point( mapping = ggplot2::aes( size = Total_new_cases ) ) +
  ggplot2::geom_smooth() +
  breaks.date +
  ggplot2::xlab( label = "Date" ) +
  ggplot2::ylab( label = "Proportion (%) of Danish Municipalities\nwith Daily New Cases of COVID-19" ) +
  ggplot2::guides( size = ggplot2::guide_legend( title = "Total Number of\nDaily New\nCOVID-19 Cases\nin Denmark" ) )

```

# Cases by Municipality

## Convert Data into Long Format

```{r}

data.plot <- 
  tidyr::pivot_longer( 
    data = data,
    cols = -1,
    names_to = "Municipality",
    values_to = "New_cases"
  )

```

## Plot

### All Municipalities

```{r Cases-by-Municipality-All, fig.height = 10, fig.width = 10, dpi = 600, dev=c( "png" )}

ggplot2::ggplot(
  data = data.plot,
  mapping =
    ggplot2::aes(
      x = date_sample,
      y = New_cases,
      color = Municipality
    )
) +
  ggplot2::geom_point( alpha = 0.25 ) +
  ggplot2::geom_smooth() +
  breaks.date +
  ggplot2::xlab( label = "Date" ) +
  ggplot2::ylab( label = "New Daily COVID-19 Cases" )

```

## Municipalities Individually

```{r Cases-by-Municipality-Individual, fig.width = 6, fig.height = 6, dpi = 600, dev=c( "png" )}

municipalities <- unique( data.plot$"Municipality" )

municipalities <- sort( x = municipalities )

for ( i in 1:length( municipalities ) ) {
  
  plot.i <-
    ggplot2::ggplot(
      data = data.plot[ data.plot$"Municipality" == municipalities[ i ], ],
      mapping =
        ggplot2::aes(
          x = date_sample,
          y = New_cases,
        )
    ) +
    ggplot2::geom_point() +
    ggplot2::geom_smooth() +
    breaks.date +
    ggplot2::xlab( label = "Date" ) +
    ggplot2::ylab( label = "Daily New COVID-19 Cases" ) +
    ggplot2::ggtitle( label = municipalities[ i ] )
  
  print( plot.i )
  print( municipalities[ i ] )
  
}

```

# SessionInfo

```{r}

utils::sessionInfo()

```

